---
title: "Lab 8 – Flint Water (Sample Solutions)"
output:
  html_document:
    df_print: paged
    toc: yes
  html_notebook:
    theme: readable
    toc: yes
    toc_float: yes
  pdf_document: default
---

### Assignment Overview

-	We’ll be investigating some data on water samples taken from Flint, Michigan in 2016, during [the water crisis](https://www.cdc.gov/nceh/casper/pdf-html/flint_water_crisis_pdf.html). We have the following analytical goals for this lab:
  + Evaluate the extent to which lead levels are unsafe across a representative sample of homes in the area
  + Investigate whether letting the faucet “flush” for a period of time lowers lead levels
  + Determine whether lead levels vary across wards (i.e., regions of the city).


### Step 0

-	Pre-lab work: Complete the pre-lab tutorials (“Customizing ggplot2” and “if else statements”) for Lab 8 first: https://stat212-learnr.stat.illinois.edu/ 

- Load `tidyverse` package.

```{r}

```

-	Download the Flint.xlsx file to your computer and then load it into your RMarkdown file. You might also need to library `readxl` to do that.


```{r message=FALSE}
library(readxl)
Flint = read_excel("Flint.xlsx")

# Note, your data file should be saved in the same folder as your RMarkdown. 
# Also match the name of the data reference above with the name of the data file. For example, it might save as Flint.xlsx, or Flint_1_.xlsx or something else.
```

- Use the following code after your import code to allow `Ward` to be read categorically. `Flint$Ward = as.character(Flint$Ward)`

```{r}
Flint$Ward = as.character(Flint$Ward)

#View(Flint)

```


### Data Description

This data set includes lead content measurements taken from tap water across 300 homes in Flint, Michigan (of which 269 homes’ measurements are included). Researchers collected 3 water samples from each household: the water at first draw (faucet turned on), water after running the faucet for 45 seconds, and water after running the faucet for 2 minutes. Lead content is measured in parts per billion (ppb). The spreadsheet is organized such that one water sample is the unit of observation; there are 3 units of observation per household.

As a point of reference, lead measurements **above 5ppb** are considered **somewhat unhealthy for regular consumption**, and lead measurements **above 15ppb** are considered **dangerous for regular consumption**.

**SampleID:** Household number. There are 269 households that provided data.

**Zip_Code:** Household’s zip code

**Ward:** The regional zone that the household was in. A ward is like a precinct.
https://www.cityofflint.com/city-of-flint-ward-map/ 

**Time:** The time point at which the water sample was taken: First draw, after 45 seconds, or after 2 minutes.

**Lead_ppb:** The lead concentration in the water sample, measured in parts per billion (ppb)


## Question 1

To get started, let’s visualize what our variable of interest looks like. Create a density curve that plots the distribution of the Lead_ppb variable. 

-	Use a fill color
-	Add a title, and label the x axis to state "Lead in parts per billion (ppb)"
-	Scale the x axis to have more frequent tick marks than is shown by default (you be the judge)

```{r}

```

**Report the numeric summary result of this variable (min, Q1, Q2, mean, Q3, max).**

```{r}

```

**Describe the shape of this distribution.**




## Question 2

Now, let’s create jitter plots to visually compare the lead levels at the three different time points. Have each group be represented as a column of jittered points.

-	Jitter your points with a width around 0.05 or 0.1
-	Set an transparency (alpha) level between around 0.2 to 0.4
-	Color each column differently, and customize the colors (manual, or with a color palette)
-	Add a title, and label the lead axis to “Lead in parts per billion (ppb)”
-	Use a plot theme


```{r}

```


## Question 3

Using a pipe, create a summary table to compare lead levels, separated in rows by the three time points of data (first draw, 45 seconds, and 2 minutes). Your summaries of lead levels should include the following.

-	The median lead level, rounded to 3 decimal places
-	The mean lead level, rounded to 3 decimal places
-	Proportion of water samples with a lead level above 5ppb, rounded to 3 decimal places
-	Proportion of water samples with a lead level above 15ppb, rounded to 3 decimal places


```{r}

```

Notice that the first draw and 45-second groups have nearly identical means. Consider your previous graph and briefly explain: **Which group do you think is producing the consistently highest lead measurements? What might be causing that spike in the mean for the other group? **


## Question 4

For sake of visualizing, let’s narrow in on where most of the data is.
Create side by side **boxplots** of these same two variables and… 

-	Color (or fill) each boxplot differently and use custom colors (or a color palette) 
-	Add a title, and label the lead axis to “Lead in parts per billion (ppb)”
-	Use the scale function to have the y axis go in increments of 5 and have **limits** from 0 to 50
-	Add a plot theme background
-	Remove the color legend (you can do this in the theme() function).

_Note the warning about “rows containing non-finite values” just means data points were excluded because they were outside the 0 to 50 range you limited your graph to—we did that on purpose!_

```{r}

```




## Question 5

Let’s now compare lead concentrations across Wards (precincts). Using a pipe, let’s create a summary table to compare lead concentrations across Wards. Filter the data to _only include_ **first-draw** _observations_. Your summaries of lead concentrations should include the following:

-	The median lead level, rounded to 3 decimal places
-	The mean lead level, rounded to 3 decimal places
-	Proportion of water samples with a lead level above 5ppb, rounded to 3 decimal places
-	Proportion of water samples with a lead level above 15ppb, rounded to 3 decimal places


```{r}

```


## Question 6

Using a pipe, filter to only include first draw measurements. Then inside this pipe, build side-by-side jitter plots to compare lead ppb (at first draw) across each ward. The reason we changed Ward to a factor variable was for this graph.

-	Color the points from each Ward a different color. Use a **color palette** for this one
-	Add a title, and label the lead axis to “Lead in parts per billion (ppb)”
-	Use the scale function to have the y axis go in increments of 5 and have **limits** from 0 to 50
-	Add a plot theme background
-	Remove the color legend (you can do this in the theme() function).


```{r}

```



## Question 7

Now, let’s make a graph that compares lead levels across Wards and across times together. Keep Ward on the x axis, and now map Time as the fill color.

- Use the boxplot geometry for this one
-	Color (or fill) each boxplot a different color. Use a color palette for this one
-	Add a title, and label the lead axis to “Lead in parts per billion (ppb)”
-	Use the scale function to have the y axis go in increments of 5 and have limits from 0 to 50
-	Add a plot theme background
-	Remove the color legend (you can do this in the theme() function).

```{r}

```

If you were advising a resident in Flint who hadn’t had their water tested for lead, **could you give them any data-based advice about how to maximize their safety if using water from their faucet?**



